Developmental trajectories of word recognition: a priming study in bilinguals

### 2023-11-22 manuscript/manuscript.qmd
New version of the interim discussions

Following @mani2010infants, each trial in the task started with the silent presentation of a prime picture. Both monolingual and bilingual infants were expected to implicitly name the prime picture. According to the language non-selective hypothesis of lexical access, bilinguals should generate two labels for the prime picture, one in each language. To test this prediction, we manipulated the phonological similarity between the prime and the target words in both languages (see @fig-task). In *Related/Non-cognate* trials, prime and target labels shared phonological onset in only the language of test. For instance an infant tested in Catalan would be presented with a chair as prime picture ($\textbf{c}\text{adira}_{\text{CAT}}/\text{silla}_{\text{SPA}}$) and with $\textbf{c}\text{ullera}_{\text{CAT}}$ [spoon] as target label. In line with @mani2011phonological, we anticipated that the phonological overlap between prime and target should modulate target word recognition in both monolinguals and bilinguals. This should be reflected in a delayed target looking preference, compared to *Unrelated* trials, in which prime and target did not share phonological onset. In *Related/Cognate* trials, the prime shared phonological onset with the target in both languages. For instance, the same infants tested in Catalan would be presented with a car as prime picture ($\textbf{c}\text{otxe}_{\text{CAT}}/\textbf{c}\text{oche}_{\text{SPA}}$) and with $\textbf{c}\text{ullera}_{\text{CAT}}$ [spoon] as target label. In bilinguals, parallel activation of the prime in both languages should increase the cohort of the target word, leading to stronger interference effects in this condition, compared to *Related/Non-cognate* and *Unrelated* trials.

![Predicted priming effects (or their absence) in the *Related/Cognate*, *Related/Non-cognate*, and *Unrelated* conditions, with examples for a participant tested in Catalan. Words represent lexical representations. Lexical representations of the task-relevant language (Catalan) are depicted inside grey boxes. Solid arrows indicate within-language priming effects, and dashed lines indicate cross-language priming effects. In *Related/Cognate* (A) and *Related/Non-cognate* (B) trials, the recognition of the Catalan target word \textipa{kuˈʎe.R@} [spoon] is predicted to be modulated by the prior activation of the prime label in Catalan. In *Related/Cognate* trials, the parallel activation of the prime label in Spanish is predicted to increase the strength of the priming effect.](_assets/img/hypotheses.png){fig-align="center"}

In line with previous studies in monolinguals, we further predicted that the strength of the lexical interference effects in the *Related/Non-cognate* and *Related/Cognate* conditions would be associated to participants' vocabulary size. Target word recognition should be delayed by the activation of a larger cohort of phonologically related words [@mani2011phonological; @chow2018phonological]. In the case of bilinguals, two measures of vocabulary size are relevant: *total vocabulary size* and *dominant-language vocabulary size*. Total vocabulary size refers to the number of words that a child has acquired, regardless of the language they belong to. Dominant-language vocabulary size (L1 vocabulary size, henceforth) refers to the number of acquired words in the dominant language of the child (in which they were tested), the language in which they know more words. If the size of interference effects are more tightly associated to bilinguals’ total vocabulary size, it would support the assumption that lexical representations from both languages enter a unique cohort in bilinguals.

Because of the short-lived effects of cross-language activation on lexical processing, and to maximise the probability of detecting priming effects, we introduced a change in the sequence of the trials relative to the original implementation by @mani2010infant. We presented target auditory labels inmediately after the offset of the prime picture, and before the onset of the target and distractor pictures. By presenting prime pictures and target auditory labels closer in time, implicit naming of the prime picture should be more likely to influence the recognition of the target word. To test the effects of this methodological change, we first run a control experiment, Study 1, in which we tested a group of same-aged English monolinguals.

@@

In this study, we conducted a conceptual replication of @mani2010infant and @mani2011phonological. We tested a group of English monolinguals aged 20 to 32 months, the same population tested in the original studies. For consistency with Study 1, we also manipulated the phonological relationship between the prime and target labels at word onset, as well as the cognate status of the prime label. Evidence from implicit naming of prime images would be revealed by differences in target looking or its time course across the trials between *Unrelated* trials and *Related/Non-cognate*, and between *Related/Non-cognate* and *Related/Cognate* trials. Given that participants in this study are monolinguals---and contrary to the monolinguals in Barcelona, have received no exposure to any other language---we should not find the cognate status of the prime labels to exert any effect on subsequent target recognition.

@@

All materials, data, and reproducible code can be found at the OSF ([https://osf.io/hy984/](https://osf.io/ckydb/)) and GitHub (<https://github.com/gongcastro/cognate-priming>) repositories. This study was conducted according to guidelines laid down in the Declaration of Helsinki, and was approved by the Drug Research Ethical Committee (CEIm) of the IMIM Parc de Salut Mar, reference 2020/9080/I. Before every testing session, caregivers were asked to read and sign an informed consent form, and were given a token of appreciation at the end of it.

@@

We collected data from `r n_participants_total` children (`r n_participants_sex$n[1]` female, `r n_participants_sex$n[2]` male, with `r english::words(n_participants_sex$n[3])` additional participants' sex not being reported; Age: *Mean* = `r round(ages$age_mean, 2)` months, *SD* = `r round(ages$age_sd, 2)`, *Range* = \[`r round(ages$age_min, 2)`-`r round(ages$age_max, 2)`\]), living in the Oxfordshire area (United Kingdom). Participants were tested at the Oxford BabyLab at the University of Oxford. Families were recruited from maternity rooms in private hospitals and social media, and contacted via phone when the child's age spanned between 20 and 32 months. From the `r n_participants_total` children that participated, `r n_participants_sessions$session_1` participated once, and `r n_participants_sessions$session_2` participated twice. Recurrent participants were tested with at least `r round(time_diffs, 2)$diff_min` months of difference. We gathered a total of `r n_sessions_total$n` testing sessions. All participants were being raised in exclusively British English monolingual homes.

```{r vocab-values-oxf}
vocabulary_oxf <- inner_join(vocabulary,
    select(participants_oxf, session_id),
    by = join_by(session_id)
)
```

We collected vocabulary data using parental responses to the Oxford Communicative Development Inventory (OCDI) [@hamilton2000infant]. Families were sent the questionnaire immediately after each experimental session, and were given two weeks to fill it. Since all vocabulary sizes of English monolingual infants were only assessed in English, both *L1 vocabulary* and *total vocabulary* sizes were identical (see @fig-vocabulary-oxf).

```{r fig-vocabulary-oxf}
#| label: fig-vocabulary-oxf
#| fig-cap: "Participant receptive vocabulary sizes across ages and language profiles. Vocabulary size scores from the recurrent participants are shown linked."
#| fig-width: 4
#| fig-height: 2.5
vocabulary_oxf |>
    inner_join(filter(attrition_participants, is_valid_participant),
        by = join_by(session_id)
    ) |>
    inner_join(select(participants, child_id, session_id, age_group),
        by = join_by(child_id, session_id)
    ) |>
    ggplot(aes(age_group, l1_prop)) +
    geom_line(aes(group = child_id),
        colour = "grey",
        alpha = 1 / 4, linewidth = 3 / 4
    ) +
    geom_dots(
        dotsize = 2,
        colour = clrs[5],
        fill = clrs[5],
        scale = 1,
        side = "both",
        alpha = 3 / 4
    ) +
    stat_summary(
        fun.data = mean_se,
        geom = "errorbar",
        linewidth = 1 / 2,
        width = 0.1,
        colour = "black",
        show.legend = FALSE
    ) +
    stat_summary(
        fun = mean,
        geom = "point",
        size = 1.5,
        colour = "black",
        show.legend = FALSE
    ) +
    # geom_point(alpha = 1) +
    labs(
        x = "Age (months)",
        y = "Vocabulary size",
        colour = "Group",
        fill = "Group"
    ) +
    scale_y_continuous(
        labels = scales::percent,
        limits = c(0, 1)
    ) +
    theme(
        panel.grid.major.y = element_line(
            linetype = "dotted",
            colour = "grey"
        ),
        legend.position = "top",
        legend.title = element_blank()
    )
```

```{r tbl-participants}
#| label: tbl-participants
#| tbl-cap: "Demographic and linguistic profile of testing sessions."
participants_bcn |>
    unnest_wider(doe) |>
    select(-matches("others")) |>
    arrange(lp, test_language) |>
    distinct(session_id, child_id, .keep_all = TRUE) |>
    summarise(
        n = n(),
        n_participants = length(unique(child_id)),
        across(age, lst(mean, sd, min, max)),
        across(matches("doe_"), lst(mean, sd)),
        # age_vctr = list(age),
        .by = c(lp, test_language)
    ) |>
    gt(
        rowname_col = "test_language",
        groupname_col = "lp"
    ) |>
    summary_rows(
        groups = c("Monolingual", "Bilingual"),
        columns = matches("age"),
        fns = list(Total ~ mean(.)),
        fmt = ~ fmt_number(., decimals = 1, use_seps = FALSE)
    ) |>
    summary_rows(
        groups = c("Monolingual", "Bilingual"),
        columns = c(n, n_participants),
        fns = list(Total ~ sum(.)),
        fmt = ~ fmt_integer(.)
    ) |>
    summary_rows(
        groups = c("Monolingual", "Bilingual"),
        columns = matches("doe"),
        fns = list(Total ~ mean(.)),
        fmt = ~ fmt_number(., decimals = 1, scale_by = 100)
    ) |>
    grand_summary_rows(
        columns = matches("age"),
        fns = list(Total ~ mean(.)),
        fmt = ~ fmt_number(., decimals = 1)
    ) |>
    grand_summary_rows(
        columns = matches("doe"),
        fns = list(Total ~ mean(.)),
        fmt = ~ fmt_number(., decimals = 1, scale_by = 100)
    ) |>
    grand_summary_rows(
        columns = c(n, n_participants),
        fns = list(Total ~ sum(.)),
        fmt = ~ fmt_integer(.)
    ) |>
    # gt_plt_dist(age_vctr, type = "histogram") |>
    fmt_number(is.numeric, decimals = 1) |>
    fmt_integer(c(n, n_participants)) |>
    fmt_number(matches("doe"), decimals = 1, scale_by = 100) |>
    cols_merge(c(age_min, age_max),
        pattern = "{1}-{2}"
    ) |>
    cols_merge(c(age_mean, age_sd),
        pattern = "{1} ({2})"
    ) |>
    cols_merge(c(doe_catalan_mean, doe_catalan_sd),
        pattern = "{1} ({2})"
    ) |>
    cols_merge(c(doe_english_mean, doe_english_sd),
        pattern = "{1} ({2})"
    ) |>
    cols_merge(c(doe_spanish_mean, doe_spanish_sd),
        pattern = "{1} ({2})"
    ) |>
    tab_spanner("Age (months)", matches("age")) |>
    tab_spanner("Catalan", matches("catalan")) |>
    tab_spanner("English", matches("english")) |>
    tab_spanner("Spanish", matches("spanish")) |>
    tab_spanner("Degree of Exposure (%)", matches("doe")) |>
    cols_label(
        n = "Sessions",
        n_participants = "Participants",
        age_mean = "M (SD)",
        age_min = "Range",
        doe_catalan_mean = "M (SD)",
        doe_english_mean = "M (SD)",
        doe_spanish_mean = "M (SD)"
    ) |>
    tab_style(
        cell_text(style = "italic"),
        cells_column_labels()
    ) |>
    tab_style(
        cell_text(style = "italic"),
        cells_column_spanners()
    )
```

```{r vocab-values}
vocabulary_bcn <- vocabulary |>
    inner_join(select(participants_bcn, session_id, lp, age_group),
        by = join_by(session_id)
    )

vocab_means <- vocabulary_bcn |>
    summarise(across(c(total_prop, l1_prop), lst(mean, sd)),
        n = n(),
        .by = c(age_group, lp)
    ) %>%
    mutate(across(matches("prop"), scales::percent)) %>%
    split(paste0("age_", gsub(" months", "", .$age_group))) |>
    map(\(x) split(x, tolower(x$lp)))

Overall, there was a substantial increase in both total and L1 vocabulary sizes associated to age (see @fig-vocabulary). Monolinguals showed higher L1 vocabulary sizes than monolinguals: they were reported to understand an average of `r vocab_means$age_21$monolingual$l1_prop_mean` (*SD* = `r vocab_means$age_21$monolingual$l1_prop_sd`) of the words at around 21 months, `r vocab_means$age_25$monolingual$l1_prop_mean` (*SD* = `r vocab_means$age_25$monolingual$l1_prop_sd`) at 25 months, and `r vocab_means$age_30$monolingual$l1_prop_mean` (*SD* = `r vocab_means$age_30$monolingual$l1_prop_sd`) at 30 months, while bilinguals were reported to understand an average of `r vocab_means$age_21$bilingual$l1_prop_mean` (*SD* = `r vocab_means$age_21$bilingual$l1_prop_sd`) of the words at around 21 months, `r vocab_means$age_25$bilingual$l1_prop_mean` (*SD* = `r vocab_means$age_25$bilingual$l1_prop_sd`) at 25 months, and `r vocab_means$age_30$bilingual$l1_prop_mean` (*SD* = `r vocab_means$age_30$bilingual$l1_prop_sd`) at 30 months. Total vocabulary sizes showed the opposite pattern, with monolinguals infants being reported to understand `r vocab_means$age_21$monolingual$total_prop_mean` (*SD* = `r vocab_means$age_21$monolingual$total_prop_sd`) of the words at around 21 months, `r vocab_means$age_25$monolingual$total_prop_mean` (*SD* = `r vocab_means$age_25$monolingual$total_prop_sd`) at 25 months, and `r vocab_means$age_30$monolingual$total_prop_mean` (*SD* = `r vocab_means$age_30$monolingual$total_prop_sd`) at 30 months, and bilinguals being reported to understand `r vocab_means$age_21$bilingual$total_prop_mean` (*SD* = `r vocab_means$age_21$bilingual$total_prop_sd`) of the words at around 21 months, `r vocab_means$age_25$bilingual$total_prop_mean` (*SD* = `r vocab_means$age_25$bilingual$total_prop_sd`) at 25 months, and `r vocab_means$age_30$bilingual$total_prop_mean` (*SD* = `r vocab_means$age_30$bilingual$total_prop_sd`) at 30 months.

```{r fig-vocabulary}
#| label: fig-vocabulary
#| fig-cap: "Participant receptive vocabulary sizes across ages and language profiles. Vocabulary size scores are presented using two measures: L1 vocabulary size (proportion of words marked as *Understands* in the vocabulary checklist of the dominant language), and total vocabulary size (proportion of words marked as *Understands* in vocabulary checklists of both dominant and non-dominant languages). Grey lines connect vocabulary size scores from the recurrent participants. For visualisation purposes, participants were group into categorical age groups at 21, 25, 30 months, the ages around which most participants were tested. Only participants included in the final data set are shown."
#| fig-width: 6
#| fig-height: 3.5
vocabulary_bcn |>
    inner_join(filter(attrition_participants, is_valid_participant),
        by = join_by(session_id)
    ) |>
    pivot_longer(c(total_prop, l1_prop),
        names_to = "measure",
        values_to = "prop",
        names_transform = \(x) gsub("_prop", "", x)
    ) |>
    mutate(
        measure = if_else(measure == "l1",
            "L1 vocabulary",
            "Total vocabulary"
        ),
        is_imputed = if_else(is_imputed, "Imputed", "Observed") |>
            factor(levels = c("Observed", "Imputed"))
    ) |>
    ggplot(aes(age_group, prop,
        colour = lp,
        fill = lp
    )) +
    facet_wrap(~measure) +
    geom_line(aes(group = child_id),
        colour = "grey",
        alpha = 1 / 4, linewidth = 3 / 4
    ) +
    geom_dots(aes(side = if_else(lp == "Monolingual", "left", "right")),
        dotsize = 2,
        stroke = 1 / 2,
        scale = 0.5,
        alpha = 3 / 4
    ) +
    stat_summary(
        fun.data = mean_se,
        geom = "errorbar",
        linewidth = 1 / 2,
        width = 0.15,
        colour = "black",
        show.legend = FALSE,
        position = position_dodge(width = 1.15)
    ) +
    stat_summary(
        fun = mean,
        geom = "point",
        size = 1.5,
        colour = "black",
        show.legend = FALSE,
        position = position_dodge(width = 1.15)
    ) +
    # geom_point(alpha = 1) +
    labs(
        x = "Age (months)",
        y = "Vocabulary size",
        colour = "Group",
        fill = "Group"
    ) +
    scale_y_continuous(
        labels = scales::percent,
        limits = c(0, 1)
    ) +
    theme(
        panel.grid.major.y = element_line(
            linetype = "dotted",
            colour = "grey"
        ),
        legend.position = "top",
        legend.title = element_blank()
    )
```

@@

New stimuli lists were generated to adapt the experimental procedure to English-learning infants. Stimuli lists were created following the same constraints as in Study 1. Auditory stimuli were recorded following the same procedure as in Study 1, produced by Southern British English native female speaker in a toddler-directed manner. Audio and picture processing were performed following the same workflow as in Study 1.

@@

Same as in Study 1, but removing the *Group* predictor from the models, given that only monolingual infants were tested.

@@

@@

### Word recognition times

>>>>

Following @mani2010infant, each trial in the task started with the silent presentation of a prime picture. Both monolingual and bilingual infants were expected to implicitly name the prime picture. According to the language non-selective hypothesis of lexical access, bilinguals should generate two labels for the prime picture, one in each language. To test this prediction, we manipulated the phonological similarity between the prime and the target words in both languages (see @fig-hypotheses). In *Related/Non-cognate* trials, prime and target labels shared phonological onset in only the language of test. For instance an infant tested in Catalan would be presented with a chair as prime picture ($\textbf{c}\text{adira}_{\text{CAT}}/\text{silla}_{\text{SPA}}$) and with $\textbf{c}\text{ullera}_{\text{CAT}}$ [spoon] as target label. In line with @mani2011phonological, we anticipated that the phonological overlap between prime and target should modulate target word recognition in both monolinguals and bilinguals. This should be reflected in a delayed target looking preference, compared to *Unrelated* trials, in which prime and target did not share phonological onset. In *Related/Cognate* trials, the prime shared phonological onset with the target in both languages. For instance, the same infants tested in Catalan would be presented with a car as prime picture ($\textbf{c}\text{otxe}_{\text{CAT}}/\textbf{c}\text{oche}_{\text{SPA}}$) and with $\textbf{c}\text{ullera}_{\text{CAT}}$ [spoon] as target label. In bilinguals, parallel activation of the prime in both languages should increase the cohort of the target word, leading to stronger interference effects in this condition, compared to *Related/Non-cognate* and *Unrelated* trials.

![Predicted priming effects (or their absence) in the *Related/Cognate*, *Related/Non-cognate*, and *Unrelated* conditions, with examples for a participant tested in Catalan. Words represent lexical representations. Lexical representations of the task-relevant language (Catalan) are depicted inside grey boxes. Solid arrows indicate within-language priming effects, and dashed lines indicate cross-language priming effects. In *Related/Cognate* (A) and *Related/Non-cognate* (B) trials, the recognition of the Catalan target word \textipa{ku"Le.R@} [spoon] is predicted to be modulated by the prior activation of the prime label in Catalan. In *Related/Cognate* trials, the parallel activation of the prime label in Spanish is predicted to increase the strength of the priming effect.](_assets/img/hypotheses.png){#fig-hypotheses fig-align="center"}

In line with previous studies in monolinguals, we further predicted that the strength of the lexical interference effects in the *Related/Non-cognate* and *Related/Cognate* conditions would be associated to participants' vocabulary size. Target word recognition should be delayed by the activation of a larger cohort of phonologically related words [@mani2011phonological; @chow2017spoken]. In the case of bilinguals, two measures of vocabulary size are relevant: *total vocabulary size* and *dominant-language vocabulary size*. Total vocabulary size refers to the number of words that a child has acquired, regardless of the language they belong to. Dominant-language vocabulary size (L1 vocabulary size, henceforth) refers to the number of acquired words in the dominant language of the child (in which they were tested), the language in which they know more words. If the size of interference effects are more tightly associated to bilinguals’ total vocabulary size, it would support the assumption that lexical representations from both languages enter a unique cohort in bilinguals.

Because of the short-lived effects of cross-language activation on lexical processing, and to maximise the probability of detecting priming effects, we introduced a change in the sequence of the trials relative to the original implementation by @mani2010infant. We presented target auditory labels immediately after the offset of the prime picture, and before the onset of the target and distractor pictures. By presenting prime pictures and target auditory labels closer in time, implicit naming of the prime picture should be more likely to influence the recognition of the target word. To test the effects of this methodological change, we first run a control experiment, Study 1, in which we tested a group of same-aged English monolinguals. In Study 2, we tested a group of monolinguals and bilinguals learning Catalan and Spanish.

@@

In this study, we conducted a conceptual replication of @mani2010infant and @mani2011phonological. We tested a group of English monolinguals aged 20 to 32 months, living in the Oxfordshire area (United Kingdom). As highlighted in the introduction, participants were tested exclusively in English, their dominant language. As in the original studies, we manipulated the phonological relationship between the prime and the target label. In line with @mani2011phonological, we predicted participants' target looking to change as a function of the phonological relatedness between the prime and target English labels. This would reveal that participants implicitly named the prime pictures, generating a phonological label that influenced the subsequent recognition of a phonologically related word. To establish a monolingual baseline for all conditions in Study 2, we also manipulated the cognate status of the prime in English and Spanish: in half of the trials the English prime label was phonologically related to its Spanish translation. Given participants' lack of familiarity with Spanish (or any language other than English), participants' performance was predicted to be unaffected by the cognate status of the primes.

@@

All materials, data, and reproducible code can be found at the OSF ([https://osf.io/hy984/](https://osf.io/ckydb/)) and GitHub (<https://github.com/gongcastro/cognate-priming>) repositories. This study was conducted according to guidelines laid down in the Declaration of Helsinki, and was approved by the Drug Research Ethical Committee (CEIm) of the IMIM Parc de Salut Mar, reference 2020/9080/I and the Medical Sciences Research Ethics Board at the University of Oxford, reference R60939/RE009. Before every testing session, caregivers were asked to read and sign an informed consent form, and were given a token of appreciation at the end of it.

@@

We collected data from `r n_participants_total` children (`r n_participants_sex$n[1]` female, `r n_participants_sex$n[2]` male, with `r english::words(n_participants_sex$n[3])` additional participants' sex not being reported; Age: *Mean* = `r round(ages$age_mean, 2)` months, *SD* = `r round(ages$age_sd, 2)`, *Range* = [`r round(ages$age_min, 2)`-`r round(ages$age_max, 2)`]) (see @tbl-participants for a detailed summary of participants' age and language profile), living in the Oxfordshire area (United Kingdom). Participants were tested at the Oxford BabyLab at the University of Oxford. Families were recruited from maternity rooms in private hospitals and social media, and contacted via phone when the child's age spanned between 20 and 32 months. From the `r n_participants_total` children that participated, `r n_participants_sessions$session_1` participated once, and `r n_participants_sessions$session_2` participated twice. Recurrent participants were tested with at least `r round(time_diffs, 2)$diff_min` months of difference. We gathered a total of `r n_sessions_total$n` testing sessions. All participants were being raised in exclusively British English monolingual homes. Participants' vision was normal, none used glasses or any other type of vision corrector.

```{r tbl-participants}
#| label: tbl-participants
#| tbl-cap: "Demographic and linguistic profile of testing sessions. The number of excluded testing sessions and participants is indicated between parentheses."
participants |>
	unnest_wider(doe) |>
	select(-matches("others")) |>
	arrange(lp, test_language) |>
	distinct(session_id, child_id, .keep_all = TRUE) |>
	mutate(study = if_else(location=="Oxford", 
						   "Study 1",
						   "Study 2"),
		   test_language = paste0(test_language, " dominant"),
		   lp = gsub(" \\(English\\)", "", lp)) |> 
	inner_join(attrition_participants,
			   by = join_by(session_id)) |> 
	summarise(n = n(),
			  n_participants = length(unique(child_id)),
			  across(age, lst(mean, sd, min, max)),
			  across(matches("doe_"), lst(mean, sd)),
			  # age_vctr = list(age),
			  .by = c(lp, study, test_language, is_valid_participant)) |> 
	pivot_wider(id_cols = c(lp, study, test_language),
				names_from = is_valid_participant,
				values_from = is.numeric) |> 
	select(lp, study, test_language, starts_with("n_"), matches("TRUE"),
		   n_FALSE, n_participants_FALSE) |> 
	rename_with(\(x) gsub("_TRUE", "", x)) |> 
	gt(rowname_col = "test_language",
	   groupname_col = c("study", "lp")) |>
	grand_summary_rows(columns = matches("age"),
					   fns = list(Total ~ mean(.)),
					   fmt = ~ fmt_number(., decimals = 1)) |>
	grand_summary_rows(columns = c(n, n_participants),
					   fns = list(Total ~ sum(.)),
					   fmt = ~ fmt_integer(.)) |>
	# gt_plt_dist(age_vctr, type = "histogram") |>
	fmt_number(is.numeric, decimals = 1) |>
	fmt_integer(c(n, n_participants, n_FALSE, n_participants_FALSE)) |>
	fmt_number(matches("doe"), decimals = 1, scale_by = 100) |>
	cols_merge(c(n, n_FALSE), pattern = "{1} ({2})") |>
	cols_merge(c(n_participants, n_participants_FALSE),
			   pattern = "{1} ({2})") |>

	cols_merge(c(age_min, age_max), pattern = "{1}-{2}") |>
	cols_merge(c(age_mean, age_sd),
			   pattern = "{1} ({2})") |>
	cols_merge(c(doe_catalan_mean, doe_catalan_sd),
			   pattern = "{1} ({2})") |>
	cols_merge(c(doe_english_mean, doe_english_sd),
			   pattern = "{1} ({2})") |>
	cols_merge(c(doe_spanish_mean, doe_spanish_sd),
			   pattern = "{1} ({2})") |>
	tab_spanner("Sample size", starts_with("n")) |>
	tab_spanner("Age (months)", matches("age")) |>
	tab_spanner("Catalan", matches("catalan")) |>
	tab_spanner("English", matches("english")) |>
	tab_spanner("Spanish", matches("spanish")) |>
	tab_spanner("Degree of Exposure (%)", matches("doe")) |>
	cols_label(n = "Sessions",
			   n_participants = "Participants",
			   age_mean = "M (SD)",
			   age_min = "Range",
			   doe_catalan_mean = "M (SD)",
			   doe_english_mean = "M (SD)",
			   doe_spanish_mean = "M (SD)") |>
	tab_style(cell_text(style = "italic"),
			  cells_column_labels()) |>
	tab_style(cell_text(style = "italic"),
			  cells_column_spanners())
```

```{r vocab-values-oxf}
vocabulary_oxf <- inner_join(vocabulary,
							 select(participants_oxf, session_id),
							 by = join_by(session_id))

vocabulary_oxf_exc <- nrow(participants_oxf) - nrow(vocabulary_oxf)
```

We collected vocabulary data using parental responses to the Oxford Communicative Development Inventory (OCDI) [@hamilton2000infant]. Families were sent the questionnaire immediately after each experimental session, and were given two weeks to fill it. In the case that a complete response to the OCDI was not provided within the two-week limit, the participants' testing session was excluded from the analyses (*n* = `r vocabulary_oxf_exc`). @fig-vocabulary-oxf shows the distribution of participants' vocabulary sizes across ages.

```{r fig-vocabulary}
#| label: fig-vocabulary-oxf
#| fig-cap: "Participant receptive vocabulary sizes across ages and language profiles. Vocabulary size scores from the recurrent participants are shown linked. For visualisation purposes, testing sessions were aggregated at three age points: 21, 25, and 30 months. Points and intervals indicate mean and standard deviations, respectively."
#| fig-width: 6
#| fig-height: 4
vocab_plot_data <- vocabulary |>
	inner_join(filter(attrition_participants, is_valid_participant),
			   by = join_by(session_id)) |>
	inner_join(select(participants, location, lp, 
					  child_id, session_id, age_group),
			   by = join_by(child_id, session_id)) |>
	pivot_longer(c(l1_prop, total_prop),
				 names_to = "type",
				 values_to = "prop") |> 
	mutate(location = factor(location, levels = c("Oxford", "Barcelona")),
		   lp = gsub(" \\(English\\)", "", lp),
		   side = case_when(location=="Oxford" ~ "both",
		   				 lp=="Monolingual" ~ "left",
		   				 lp=="Bilingual" ~ "right"),
		   type = if_else(type=="l1_prop", 
		   			   "L1 vocabulary", "Total vocabulary"),
		   study = if_else(location=="Oxford", 
		   				"Study 1 (English)", "Study 2 (Catalan/Spanish)"))

vocab_plot_means <- vocab_plot_data |> 
	summarise(prop_sd = sd(prop, na.rm = TRUE),
			  prop = mean(prop, na.rm = TRUE),
			  n = n(),
			  .by = c(age_group, type, side, lp, type, location)) |> 
	mutate(.lower = prop - (prop_sd/sqrt(n)),
		   .upper = prop + (prop_sd/sqrt(n)))

vocab_plot_data |> 
	ggplot(aes(age_group, prop, 
			   colour = lp,
			   fill = lp,
			   side = side)) +
	facet_grid(type~study) +
	geom_line(aes(group = child_id),
			  colour = "grey",
			  alpha = 1 / 4, linewidth = 3 / 4) +
	geom_dots(aes(side =),
			  dotsize = 1.5,
			  scale = 0.25,
			  binwidth = 0.025,
			  alpha = 2 / 4) +
	geom_errorbar(data = vocab_plot_means,
				  aes(ymin = .lower,
				  	ymax = .upper),
				  colour = "black",
				  show.legend = FALSE,
				  width = 0.15,
				  linewidth = 1/2,
				  position = position_dodge(width = 1)) +
	geom_point(data = vocab_plot_means, 
			   colour = "black",
			   show.legend = FALSE,
			   size = 1.25,
			   position = position_dodge(width = 1)) +
	# geom_label(data = vocab_plot_means,
	# 		   aes(label = sprintf("%.1f%%\n[%.1f, %.1f]",
	# 		   					prop*100, .lower*100, .upper*100),
	# 		   	y = 0),
	# 		   size = 2.5,
	# 		   fill = "white",
	# 		   vjust = 0,
	# 		   show.legend = FALSE,
	# 		   label.r = unit(0, "lines"),
	# 		   label.size = 0,
	# 		   label.padding = unit(0.1, "lines"),
# 		   position = position_dodge(1),
# 		   alpha = 1/4) +
# geom_point(alpha = 1) +
labs(x = "Age (months)",
	 y = "Vocabulary size",
	 colour = "Group",
	 fill = "Group") +
	scale_y_continuous(labels = scales::percent,
					   limits = c(0, 1)) +
	theme(panel.grid.major.y = element_line(linetype = "dotted",
											colour = "grey"),
		  legend.position = "top",
		  axis.title.x = element_blank(),
		  legend.title = element_blank())
```

@@

Participants were presented with 32 trials in random order, which belonged to three conditions: *Unrelated* (16 trials), *Related/Non-cognate* (8 trials), and *Related/Cognate* (8 trials). In *Unrelated* trials, target was phonologically unrelated with the English and Spanish labels of the prime picture (e.g., $\text{door}_{\text{ENG}}/\text{puerta}_{\text{SPA}}-\text{sock}_{\text{ENG}}$). In *Related/Non-cognate* trials, the target shared phonological onset with the English prime label, but not with the Spanish prime label (e.g., $\textbf{t}\text{ree}_{\text{ENG}}/\text{árbol}_{\text{SPA}}-\textbf{t}\text{ruck}_{\text{ENG}}$). In *Related/Cognate* trials, the target shared phonological overlap with both English and Spanish prime labels ($\textbf{b}\text{oot}_{\text{ENG}}/\textbf{b}\text{ota}_{\text{SPA}}-\textbf{b}\text{ird}_{\text{ENG}}$). Target-distractor pairs were phonologically unrelated, and held constant for all participants. Especial attention was paid to avoiding semantic or taxonomic relationships between prime and target words, and between prime and distractor words. Each target distractor pair was shown twice for each participant, switching roles. Distractors were always phonologically unrelated to the prime and target labels in the same trial.

We created fours lists of trials, across which the same target-distractor pair appeared with a different prime, counterbalancing the condition to which it belonged. For instance, in list one the *ball*-*trousers* pair was preceded by *bike* (*Related/Cognate*), by *butterfly* (*Related/Non-cognate*) in list two, and by *star* and *nose* (*Unrelated*) in lists three and four. @tbl-stimuli shows a detailed summary of the stimuli properties, broken down by trial type and testing language. Lexical frequencies were extracted from the English corpora from the CHILDES database [@macwhinney2000childes; @sanchez2019childes] as counts per million words, and transformed into Zipf scores for easier cross-language comparison [@zipf1945meaning; @van2014subtlex].

```{r tbl-stimuli}
#| label: tbl-stimuli
#| tbl-cap: "Summary of stimuli properties by trial type."
stimuli |>
	unnest_wider(c(freq, nphon)) |>
	summarise(across(c(matches("n_phon|freq_|lv_|nphon"), duration),
					 list(	mean = \(x) mean(x, na.rm = TRUE),
					 	  sd = \(x) sd(x, na.rm = TRUE))),
			  .by = c(trial_type, test_language)) |>
	select(-matches("distractor")) |>
	mutate(test_language = if_else(test_language=="English",
								   paste0("Study 1: ", test_language),
								   paste0("Study 2: ", test_language))) |>
	arrange(test_language) |> 
	gt(rowname_col = "trial_type",
	   groupname_col = "test_language") |>
	# summary_rows(groups = c("Unrelated", "Non-cognate", "Cognate"),
	# 			 columns = is.numeric,
	# 			 fns = list(Total ~ mean(.)),
	# 			 fmt = ~ fmt_number(., decimals = 2)) |>
	grand_summary_rows(	columns = is.numeric,
						fns = list(Total ~ mean(.)),
						fmt = ~ fmt_number(., decimals = 2)) |>
	tab_spanner("Phonemes", matches("nphon")) |>
	tab_spanner("Frequency (Zipf)", matches("freq")) |>
	tab_spanner("Phon. similarity [0-1]", matches("lv")) |>
	fmt_number(is.numeric, decimals = 1) |>
	cols_merge(c(freq_prime_mean, freq_prime_sd),
			   pattern = "{1} ({2})") |>
	cols_merge(c(freq_target_mean, freq_target_sd),
			   pattern = "{1} ({2})") |>
	cols_merge(c(nphon_prime_mean, nphon_prime_sd),
			   pattern = "{1} ({2})") |>
	cols_merge(c(nphon_target_mean, nphon_target_sd),
			   pattern = "{1} ({2})") |>
	cols_merge(c(lv_pp_mean, lv_pp_sd),
			   pattern = "{1} ({2})") |>
	cols_merge(c(lv_pt_mean, lv_pt_sd),
			   pattern = "{1} ({2})") |>
	cols_merge(c(duration_mean, duration_sd),
			   pattern = "{1} ({2})") |>
	cols_label(trial_type = "",
			   freq_prime_mean = "Prime",
			   freq_target_mean = "Target",
			   lv_pp_mean = "Prime-Prime",
			   lv_pt_mean = "Prime-Target",
			   duration_mean = "Duration (s)",
			   nphon_prime_mean = "Prime",
			   nphon_target_mean = "Target") |>
	tab_style(cell_text(style = "italic"),
			  list(cells_column_labels(),
			  	 cells_column_spanners(),
			  	 cells_grand_summary()))
```

The auditory stimuli were natural exemplars of the selected target words, spoken by a Southern British English female speaker who was instructed to pronounce each word in a toddler-directed manner. We used the Audacity and Praat [@boersma2001speak] software packages to trim, denoised, and normalise their amplitude. The visual stimuli were realistic photographic representations of a typical exemplars of the prime, target, and distractor words. Image backgrounds were removed from the original pictures using the GNU Image Manipulation Program (GIMP), resized to a rectangle of a maximum of 400 pixels hight or wide, and finally placed in the centre of a 50% grey rectangle square of 500 $\times$ 500 pixels. The final stimuli had a resolution of 72 dpi. When presented in the eye-tracker screen, the areas of interest (AOI) occupied an area of 13.23 $\times$ 13.23 cm (11.613$^{\circ}$ visual angle from participants' perspective).

@@

@@

Due to time constraints induced by the COVID-19 lockdown between 2020 an 2021, data collection in Study 1 and Study 2 was conducted simultaneously. In this study, we present data for monolinguals and bilinguals tested in Barcelona.

@@
